Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

### 2021-03-11 doc/SA-Hermaphrodites-wDemography.tex
Expanded Discussion, New Abstract, Table, etc.
Expanded Discussion, added a couple alternative abstracts, finished
(hopefully) table of model parameters, and made minor edits
throughout the ms.

Still have not updated appendices. Need to chat about these I think.

\tableofcontents
%%%%%%%%%%%%%%%%%%%%
\newpage{}
\section{Abstract}
%%%%%%%%%%%%%%%%%%%%
\ldots

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%

Population persistence or extinction depends heavily on the ability of a species to adapt to the environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1957}, environmental shifts \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, or genetic constraints \citep{ConnallonHall2018}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when there are genetic trade-offs between fitness components \citep{CharlesworthHughes2000, ConnallonHall2018}.

 % Whether a population persists or goes extinct depends in large part on its ability to adapt to the environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, including deleterious mutations \citep{Haldane1957}, environmental shifts causing a population to chase a new phenotypic optimum \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, as well as genetic constraints arising from intrinsic features of the inheritance system \citep{ConnallonHall2018}.

%Conflicting selection and gene-flow between different classes of individuals, or genetic trade-offs between fitness components, can impose constraints on adaptation with particularly interesting evolutionary consequences \citep{CharlesworthHughes2000, ConnallonHall2018}. 
Fitness trade-offs constraining adaptation have particularly interesting evolutionary consequences. On the one hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase the population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. On the other hand, they provide an effective mechanism for the maintenance of genetic variation, which influences the population's capactiy for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the contribution of genetic constraints to the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection, due to genetic trade-offs between male and female fitness, is a common feature of sexually reproducing populations, and is thought to contribute not only to the maintenance of genetic variation, but also to the evolution of sexual dimorphism and gender-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, "sexually antagonistic" selection (abbreviated SA hereafter) can arise whereby beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}. Alleles with opposing fitness effects through male and female sex functions can cause analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}.

The evolutionary dynamics of sexually antagonistic alleles in hermaphrodites differ from those in dioecious populations because individuals may reproduce through a combination of self- and outcross fertilization \citep{Goodwillie2005, Igic2006, JarneAuld2006}. For example, self-fertilization is predicted to reduce the total parameter space where balanced SA polymorphisms can be maintained, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014}; but see \citealt{Tazzyman2015}). Yet, other processes, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, or spatial heterogeneity and the complexity of the life-cycle can expand the parameter space for SA polymorphism \citep{Olito-etal-2018,ConnallonSharmaOlito2019}. Overall, both theoretical predictions and current empirical data suggest there is ample scope for SA trade-offs and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, WangBarrett2020}, although identifying specific SA loci from genome sequence data remains challenging \citep{RuzickaESEB2020}.

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes, and therefore do not consider the population dynamical consequences of evolutionary change. To take the population dynamical consequences into account, a model is required that links changes in genotype distributions to population dynamics.  \cite{harts2014demography} build such a model using an individual based approach. Their model tracks both the population dynamics and the spread of sexually antagonistic genes as well as dispersal genes in a species with non-overlapping generations. \cite{deVriesCaswell2019b} introduced a model with overlapping generations, intralocus sexual antagonism, and population dynamics but did not perform analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

In this paper, we extend the model by \cite{deVriesCaswell2019b} to include both dioecious and hermaphroditic species, and we use this model to study when the parameter conditions predicted to be polymorphic by population genetics models are also demographically viable. A major strength of the model is that we can use empirically estimated demographic rates to parameterize the model, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

%Such an eco-evolutionary model requires a map from genotype to phenotype, from phenotype to individual demographic rates (birth and death rates), and finally from demographic rates to population level characteristics \citep{coulson2006putting}.    
%Genetic trade-offs between fitness components due to pleiotropy represent a form of genetic constraint with particularly interesting evolutionary consequences because they act as a double-edged sword

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Model}
%%%%%%%%%%%%%%%%%%%%%%%%

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of simultaneous hermaphrodites with outcrossing or mixed-mating systems. The derivation and some of the key results follow closely those presented in \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b}; and in fact, the model presented here reduces to the two-sex stage-structured model of \citet{deVriesCaswell2019b} under obligate outcrossing. A full derivation of the model and analyses is presented in \hl{Appendix A}, and all simulation code necessary to reproduce the results are available at \url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

Simultaneous hermaphrodites express both male and female sex functions (i.e., can be both a mother and father), and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female function, we denote matrices or vectors relating to male sex function with a prime.

Individuals in the model are jointly classified by life-cycle stage ($1, \ldots, \omega$), genotype ($1, \ldots, g$), and whether they were produced by self- or outcross fertilization (denoted by $S$ and $X$ superscripts, respectively). Whether individuals were produced through selfing or outcrossing is included in the individual state description because individuals produced through selfing might experience reduced survival, growth, or maturation rates as a consequence of inbreeding depression. To model inbreeding depression, we therefore keep track of how individuals were produced. Of course the level of inbreeding depression will not only be a function of whether the current individual is produced through inbreeding but also of how many generations of inbreeding have occurred (see for example, \cite{kelly1999response} and \cite{kelly2007mutation}) but keeping track of the history of inbreeding is beyond the scope of this paper. Therefore we use the commonly made assumption that individuals produced through selfing suffer a fixed level of inbreeding depression (\textbf{which reference(s) would work best here?}).

% Each genotype is characterized by a matrix of transition probabilities (including survival), and a matrix of reproductive output through male and female function. These matrices can include time variation or nonlinearities reflecting the environment or density dependence (see for example, \cite{de2020matrix}), but these will not be modelled here. Each stage contributes offspring to genotypes at the next time step according to transition matrices that are determined by the mating system and stage-structure of the population.

@@

\noindent where $\| \cdot \|$ is the one-norm. The population vector $\tilde{\mbf{n}}(t)$ is projected forward from time $t$ to $t + 1$ by the $projection$ matrix $\tilde{\mbf{A}}(\tilde{\mbf{n}})$ such that 
\begin{linenomath*}
\begin{equation}
	\tilde{\mbf{n}}(t + 1) = \tilde{\mbf{A}}[\tilde{\mbf{n}}(t)] \, \tilde{\mbf{n}}(t)
\end{equation}
\end{linenomath*}

The population projection matrix $\tilde{\mbf{A}}$ is constructed from four sets of matrices representing the demographic and genetic processes: The matrices $\mbf{U}^{S}_{i}$ and $\mbf{U}^{X}_{i}$ contain transition and survival probabilities for each genotype, produced by selfing and outcrossing respectively. The matrices $\mbf{F}_{i}$ and $\mbf{F}^{\prime}_{i}$ contain the genotype $\times$ stage specific contribution via female and male sex-functions of genotype $i$ to the female and male gamete pools, respectively, and therefore to zygotes in the next generation. We assume that whether individuals were produced through selfing or outcrossing does not affect their fecundity or mating success, that is, we assume  $\mbf{F}^X_{i}=\mbf{F}^S_{i}=\mbf{F}_{i}$ and $\mbf{F}^{\prime X}_{i}=\mbf{F}^{\prime S}_{i}=\mbf{F}^{\prime}_{i}$. Deviations from this assumption are straightforward to incorporate but beyond the scope of the current paper.

@@

Since we assume that how individuals were produced does not affect their relative mating success, the frequency in the male mating population is obtained simply by summing the vectors of individuals produced through selfing and through outcrossing, that is, 
\begin{equation}
 	\mbf{p}^{\prime}=\frac{(\mbf{n}^X + \mbf{n}^S)}{\| (\mbf{n}^X + \mbf{n}^S)\|}.
 \end{equation} 
Depending on their stage and genotype at the SA locus, individuals contribute male gametes to the overall population gamete pool according to the following equation
\begin{linenomath*}
\begin{equation}
		\mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime}= 
		\left(
			\begin{array}{ccc}
				\mbf{1}^{\intercal}_{\omega} & \frac{1}{2} \mbf{1}^{\intercal}_{\omega} & 0 \\
				0 & \frac{1}{2} \mbf{1}^{\intercal}_{\omega} & \mbf{1}^{\intercal}_{\omega} \\
			\end{array} \right)
		\left(
			\begin{array}{ccc}
				\mbf{F}^{\prime}_{AA} & 0 & 0 \\
				0 & \mbf{F}^{\prime}_{Aa} & 0 \\
				0 & 0 & \mbf{F}^{\prime}_{aa} \\
			\end{array} \right)
		\left(
			\begin{array}{c}
				\mbf{p}^{\prime}_{AA} \\
				\mbf{p}^{\prime}_{Aa} \\
				\mbf{p}^{\prime}_{aa} \\
			\end{array} \right).
\end{equation}
\end{linenomath*}

\noindent The matrix $\mbb{F}^{\prime}$ is composed of genotype-specific fertility matrices, and operates on the the vector of genotype frequencies to give their relative contribution of each genotype to the gamete pool. The matrix $\mbf{W}^{\prime}$ converts these contributions to allele numbers. Normalizing the resulting vector gives the allele frequencies in the male gamete pool:
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool}
	\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) = 
			\frac{\mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime}}{\| \mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime} \|} = 
				\frac{\mbf{W}^{\prime} \mbb{F}^{\prime} (\mbf{n}^X + \mbf{n}^S)}{\| \mbf{W}^{\prime} \mbb{F}^{\prime} (\mbf{n}^X + \mbf{n}^S)\|}
\end{equation}
\end{linenomath*}

\textcolor{red}{
I'm doubting whether to move the next paragraph to the appendices as well.  Equation (\ref{eq:maleGametePool}) is all we strictly need. If we keep the parent-offspring maps, we can explain in the next section how the matrices 
$\mcal{F}^S$ and $\mcal{F}^X$ have the same structure. Would that be helpful? Or would it be easier to just explain in the next section what the structure of $\mcal{F}^S$ and $\mcal{F}^X$ is? }

@@

Reproduction via self-fertilization differs from outcrossing. As in previous models of partial selfing, we assume that individuals produce enough male gametes to easily self-fertilize all of their ovules, and that self-fertilization involves little or no selection from external factors relative to outcrossing (e.g., \citealt{Charlesworth1978a,JordanConnallon2014,Olito2017}, but see \citealt{Tazzyman2015}). Under these assumptions, the genotype distributions of selfed offspring are determined entirely by the parental genotype and the probabilities of segregation and fertilization during and after meiosis: 
\begin{linenomath*}
\begin{equation} \label{eq:HS}
	\mbf{H}^S(\tilde{\mbf{n}}) = 
			\left(
			\begin{array}{ccc}
				1 & 1/4 & 0 \\
				0 & 1/2 & 0 \\
				0 & 1/4 & 1 \\
			\end{array} \right).
\end{equation}
\end{linenomath*}
\noindent The columns of $\mbf{H}^S(\tilde{\mbf{n}})$ again give the selfed offspring genotype distributions for parental genotypes of $AA$, $Aa$, and $aa$, respectively.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sexually antagonistic selection \& inbreeding depression} \label{sec:SAsel}

We now construct and analyze a genotype $\times$ stage-classified model for a hypothetical species with intralocus sexual conflict via the two sex functions. To keep the model analytically tractable, we assume our hypothetical species has two life stages: juveniles and adults (i.e., $\omega = 2$), and that only adults are reproductively active. Suppose that there is a genetic trade-off between the sex-functions at a single bi-allelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Analyses} \label{subsec:analyses}

@@

We identify conditions where SA polymorphism is 'protected' by evaluating the stability of populations initially fixed for either SA allele to invasion by the other (i.e., stability was assessed at the boundary equilibrium genotype frequencies of $\hat{\mbf{p}}_{AA} = 1$ and $\hat{\mbf{p}}_{aa} = 1$; \citealt{Levene1953, Prout1968, deVriesCaswell2019b}). The formal conditions for a protected polymorphism are determined by linearizing the model in the vicinity of the boundary equilibria ($\hat{\mbf{p}}$), and evaluating the magnitude of the largest eigenvalue of the Jacobian matrix of the linearization. A full derivation of the Jacobian and details of the invasion analysis are provided in \hl{Online Appendix X.x}.

@@

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (i.e., $s < 0.1$; e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For a given parameter set, we numerically evaluated the largest eigenvalue of the Jacobian evaluated at both boundaries, and the population growth rate of the equilibrium obtained after perturbation of the boundary, $\lambda$. To determine the viability threshold (where the population growth rate equals one) and the boundary stability thresholds (where the largest eigenvalue of the Jacobian crosses one), we start at the outer boundaries of selection parameter space, and titrated inward. The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds

>>>>

%\tableofcontents
%%%%%%%%%%%%%%%%%%%%
\newpage{}
\section*{Abstract -- currently 215/200 word max.}
%%%%%%%%%%%%%%%%%%%%
Genetic trade-offs between male and female fitness impose fascinating genetic constraints on adaptation. On one hand they can maintain genetic variation for a variety of life-history and gender-related traits, while on the other hand they cause unavoidable maladaptation as individuals are prevented from reaching their sex-specific fitness optima. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of sexually antagonistic (SA) selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focus on identifying the parameter conditions under which SA polymorphisms are maintained by balancing selection {\itshape and} the population intrinsic growth rate remains positive. We also analyze the effects of inbreeding depression manifesting in different life-history stages. Finally, we gave an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results indicate that when population growth rates are near one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

\newpage{}
\section*{ALTERNATIVE Abstract -- currently 200/200 word max.}
% Alternative opening sentence to the abstract:sentence
Now classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focus on identifying the parameter conditions under which SA polymorphisms are maintained by balancing selection and the population intrinsic growth rate remains positive. We also analyze the effects of inbreeding depression manifesting in different life-history stages, and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant Mimulus guttatus. Our results indicate that when population growth rates are near one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with M. guttatus appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

% GOOD STUFF HERE... KEEPING THIS IN CASE WE WANT TO WORK MORE OF THIS INTO THE ABSTRACT...In this paper we developed a stage-structured mendelian matrix model that accommodates selection through male and female sex-functions in populations of simultaneous hermaphrodites with mixed mating systems. Using the model, we jointly analyzed the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focused our analyses on identifying the parameter conditions under which SA polymorphisms are maintained by balancing selection {\itshape and} the population is also demographically viable (i.e., has a positive intrinsic growth rate). We also looked at the effects of inbreeding depression manifesting in different life-history stages, exploring the fitness consequences of early- vs.~late-acting inbreeding depression on the model predictions. Finally, we gave an illustrative example of the potential for SA polymorphism in real populations by parameterizing our model with empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%

In order to persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1957}, changing environmental conditions \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when there are genetic trade-offs between fitness components \citep{CharlesworthHughes2000, ConnallonHall2018}.

Fitness trade-offs that impose genetic constraints on adaptation have particularly interesting evolutionary consequences. On the one hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. On the other hand, they provide an effective mechanism for the maintenance of genetic variation, which influences a population's capactiy for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) arising from genetic trade-offs between male and female fitness, is a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and also the evolution of sexual dimorphism and gender-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, SA selection can arise when beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}. Alleles with opposing fitness effects through male and female sex functions can cause analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}.

The evolutionary dynamics of sexually antagonistic alleles in hermaphrodites differ from those in dioecious populations because individuals may reproduce through a combination of self- and outcross fertilization \citep{Goodwillie2005, Igic2006, JarneAuld2006}. For example, self-fertilization is predicted to reduce the total parameter space where balanced SA polymorphisms can be maintained, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014}; but see \citealt{Tazzyman2015}). Yet, other factors, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, or spatial heterogeneity and complexity of the life-cycle can expand the parameter space for SA polymorphism \citep{Olito-etal-2018,ConnallonSharmaOlito2019}. Overall, both theoretical predictions and current empirical data suggest there is ample scope for SA trade-offs and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, WangBarrett2020}, although identifying specific SA loci from genome sequence data remains challenging \citep{RuzickaESEB2020}.

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant size, and therefore do not consider the population dynamical consequences of evolutionary change. In fact, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals. To take the population dynamical consequences of polymorphism into account, a model is required that links changes in genotype distributions to population dynamics. For example, \cite{harts2014demography} developed an individual-based modelling approach that tracks population dynamics and the spread of both SA genes and dispersal genes in a species with non-overlapping generations. \cite{deVriesCaswell2019b} introduced a mendelian matrix model with overlapping generations, intralocus sexual antagonism, and population dynamics but did not perform analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

In this paper, we extend the model of \cite{deVriesCaswell2019b} to include both dioecious and hermaphroditic species in order to study the parameter conditions under which balanced SA polymorphisms are also demographically viable. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{The Model}
%%%%%%%%%%%%%%%%%%%%%%%%

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of simultaneous hermaphrodites with outcrossing or mixed-mating systems. The derivation and some of the key results follow closely those presented in \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b}. In fact, the present model reduces to the two-sex stage-structured model of \citet{deVriesCaswell2019b} under obligate outcrossing. A full derivation of the model and analyses is presented in \hl{Appendix A}, and all computer code necessary to reproduce the results are available at \url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

Simultaneous hermaphrodites express both male and female sex functions (i.e., can transmit genes to the next generation via both sperm and ovules), and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from another individual in the population, while paternal outcrossing involves exporting male gametes to another individual and fertilizing their ovules. Self-fertilization is achieved when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female function, we denote matrices or vectors relating to male sex function with a prime.

Individuals in the model are jointly classified by life-cycle stage ($1, \ldots, \omega$), genotype ($1, \ldots, g$), and whether they were produced by self- or outcross fertilization (denoted by $S$ and $X$ superscripts, respectively). Whether individuals were produced through selfing or outcrossing is included in the individual state description because individuals produced through selfing might experience reduced survival, growth, or maturation rates as a consequence of inbreeding depression. In reality, the severity of inbreeding depression will not only be a function of whether an individual is produced by selfing but also of how many consecutive generations of inbreeding have occurred in their lineage (e.g., \citealt{kelly1999response, kelly2007mutation}). However, tracking lineage-specific inbreeding histories is beyond the scope of this paper, and we therefore make the common assumption that the severity of inbreeding depression is the same for all individuals (e.g., \citealt{Charlesworth1987, Charlesworth2009, Charlesworth2010,JordanConnallon2014}).

@@

\noindent where $\| \cdot \|$ is the one-norm. The population vector $\tilde{\mbf{n}}(t)$ is projected forward from time $t$ to $t + 1$ by the projection matrix $\tilde{\mbf{A}}(\tilde{\mbf{n}})$ such that 
\begin{linenomath*}
\begin{equation}
	\tilde{\mbf{n}}(t + 1) = \tilde{\mbf{A}}[\tilde{\mbf{n}}(t)] \, \tilde{\mbf{n}}(t)
\end{equation}
\end{linenomath*}

The population projection matrix $\tilde{\mbf{A}}$ is constructed from four sets of matrices representing the demographic and genetic processes: The matrices $\mbf{U}^{S}_{i}$ and $\mbf{U}^{X}_{i}$ contain transition and survival probabilities for each genotype, produced by selfing and outcrossing respectively. The matrices $\mbf{F}_{i}$ and $\mbf{F}^{\prime}_{i}$ contain the genotype $\times$ stage specific contribution via female and male sex-functions of genotype $i$ to the female and male gamete pools, respectively, and therefore to zygotes in the next generation. We assume that whether individuals were produced through selfing or outcrossing does not affect their fecundity or mating success, that is, we assume  $\mbf{F}^X_{i}=\mbf{F}^S_{i}=\mbf{F}_{i}$ and $\mbf{F}^{\prime X}_{i}=\mbf{F}^{\prime S}_{i}=\mbf{F}^{\prime}_{i}$. Deviations from this assumption are straightforward to incorporate but beyond the scope of this paper.

@@

Because we assume that how individuals were produced (either by selfing or outcrossing) does not affect their relative mating success, the frequency in the male mating population is obtained by simply summing the vectors of individuals produced through selfing and outcrossing:
\begin{linenomath*}
\begin{equation}
 	\mbf{p}^{\prime}=\frac{(\mbf{n}^X + \mbf{n}^S)}{\| (\mbf{n}^X + \mbf{n}^S)\|}.
 \end{equation}
\end{linenomath*}
Depending on their stage and genotype at the SA locus, individuals contribute male gametes to the overall population gamete pool according to the following equation
\begin{linenomath*}
\begin{equation}
		\mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime}= 
		\left(
			\begin{array}{ccc}
				\mbf{1}^{\intercal}_{\omega} & \frac{1}{2} \mbf{1}^{\intercal}_{\omega} & 0 \\
				0 & \frac{1}{2} \mbf{1}^{\intercal}_{\omega} & \mbf{1}^{\intercal}_{\omega} \\
			\end{array} \right)
		\left(
			\begin{array}{ccc}
				\mbf{F}^{\prime}_{AA} & 0 & 0 \\
				0 & \mbf{F}^{\prime}_{Aa} & 0 \\
				0 & 0 & \mbf{F}^{\prime}_{aa} \\
			\end{array} \right)
		\left(
			\begin{array}{c}
				\mbf{p}^{\prime}_{AA} \\
				\mbf{p}^{\prime}_{Aa} \\
				\mbf{p}^{\prime}_{aa} \\
			\end{array} \right).
\end{equation}
\end{linenomath*}

\noindent The matrix $\mbb{F}^{\prime}$ is composed of genotype-specific fertility matrices, and operates on the the vector of genotype frequencies to give their relative contribution of each genotype to the gamete pool. The matrix $\mbf{W}^{\prime}$ converts these contributions to allele numbers. Normalizing the resulting vector gives the allele frequencies in the male gamete pool:
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool}
	\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) = 
			\frac{\mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime}}{\| \mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime} \|} = 
				\frac{\mbf{W}^{\prime} \mbb{F}^{\prime} (\mbf{n}^X + \mbf{n}^S)}{\| \mbf{W}^{\prime} \mbb{F}^{\prime} (\mbf{n}^X + \mbf{n}^S)\|}
\end{equation}
\end{linenomath*}
The key difference between selfing and outcrossing reproduction can be seen in the parent-offspring maps, which reflect the joint effects of meiosis and mating on the distribution of offspring genotypes.
% \textcolor{red}{
% I'm doubting whether to move the next paragraph to the appendices as well.  Equation (\ref{eq:maleGametePool}) is all we strictly need. If we keep the parent-offspring maps, we can explain in the next section how the matrices $\mcal{F}^S$ and $\mcal{F}^X$ have the same structure. Would that be helpful? Or would it be easier to just explain in the next section what the structure of $\mcal{F}^S$ and $\mcal{F}^X$ is? }

@@

Reproduction via self-fertilization differs from outcrossing. As in previous models of partial selfing, we assume that individuals produce enough male gametes to easily self-fertilize all of their ovules, and that self-fertilization involves little or no selection from external factors relative to outcrossing (e.g., \citealt{Charlesworth1978a,JordanConnallon2014,Olito2017}, but see \citealt{Tazzyman2015}). Under these assumptions, the genotype distributions of selfed offspring are determined entirely by the parental genotype and the probabilities of segregation and fertilization during and after meiosis: 
\begin{linenomath*}
\begin{equation} \label{eq:HS}
	\mbf{H}^S(\tilde{\mbf{n}}) = 
			\left(
			\begin{array}{ccc}
				1 & 1/4 & 0 \\
				0 & 1/2 & 0 \\
				0 & 1/4 & 1 \\
			\end{array} \right).
\end{equation}
\end{linenomath*}
\noindent The columns of $\mbf{H}^S(\tilde{\mbf{n}})$ again give the selfed offspring genotype distributions for parental genotypes of $AA$, $Aa$, and $aa$, respectively. As described in the next section, these parent-offspring maps influence the projection matrix by altering the fertilitiy matrices for individuals produced by selfing and outcrossing (see Eq(\ref{eq:BlkFS}) and Eq(\ref{eq:BlkFX}) below).

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Sexually antagonistic selection \& inbreeding depression} \label{sec:SAsel}

We now construct and analyze a genotype $\times$ stage-classified model for a hypothetical species with intralocus sexual conflict via the two sex functions. To keep the model analytically tractable, we assume our hypothetical species has two life stages: juveniles and adults (i.e., $\omega = 2$), and that only adults are reproductively active. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Analyses} \label{subsec:analyses}

@@

We identify conditions where SA polymorphism is 'protected' by evaluating the stability of populations initially fixed for either SA allele to invasion by the other (i.e., we assessed stability at the boundary equilibrium genotype frequencies of $\hat{\mbf{p}}_{AA} = 1$ and $\hat{\mbf{p}}_{aa} = 1$; \citealt{Levene1953, Prout1968, deVriesCaswell2019b}). The formal conditions for a protected polymorphism are determined by linearizing the model in the vicinity of the boundary equilibria ($\hat{\mbf{p}}$), and evaluating the magnitude of the largest eigenvalue of the Jacobian matrix of the linearization. A full derivation of the Jacobian and details of the invasion analysis are provided in \hl{Online Appendix X.x}.

@@

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (i.e., $s < 0.1$; e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For simplicity, we also assume equal dominance in all analyses (i.e., $h_f = h_m = h$). For a given parameter set, we numerically evaluated the largest eigenvalue of the Jacobian evaluated at both boundaries, and the population growth rate of the equilibrium obtained after perturbation of the boundary, $\lambda$. To determine the viability threshold (where the population growth rate equals one) and the boundary stability thresholds (where the largest eigenvalue of the Jacobian crosses one), we start at the outer boundaries of selection parameter space, and titrated inward. The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds
